Insight Into Lumbar Back Pain \subtitle{What the Lumbar Spine Tells About Your Life}

### 2014-08-27 draft.tex
added draft of proposed-features figure
Revised centerline and age-gender figure

\subsection{Image Data}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.5\textwidth]{figures/centerline}
  \caption{
	(a) left: Tetrahedron-based finite element model (FEM) of the lumbar spine, capturing the L1-L5 vertebrae as well as the lumbar spine canal.
	%%
	(a) right: FEM aligned to a MRI data set.
	%%
	The purple dashed line represents the centerline, describing the lumbar spine canal with 92 points.
	%%
	(b) Using the Frenet formulas, we extract the weighted sum of curvature and torsion for all 92 points (white dashes).
	%%
	The curvature angle is calculated using the top, bottom and middle point.
	%%
	We also extracted the variables for each projection axis which to assess their information gain.
	%(b) Using the Frenet formulas, we extract the curvature, torsion and the curvature angle from the lumbar spine canal center line.
	}
  \label{fig:centerline}
\end{figure}
\com{Image Data}
Four trained technicians acquired the medical image data for each subject in a standardized way using a 1.5~Tesla MRI scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
%%
The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
%%

@@

We have to assess the model accuracy to extract key figures from it.
%%
The detection model depicts the vertebrae positions, but lacks detailed information about their volume.
%%
It captures reliable information about spine canal curvature.
%%
In a previous work \cite{Klemm2013VMV,Klemm2014VIS}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
%%
Using the Frenet formulas \cite{Frenet}, we extracted the following metrics from the model (Fig.~\ref{fig:centerline}~b):
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum \frac{curvature_i}{curvature_{all}}$.
%%
	\item \emph{Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_{all}}$.
	\item \emph{Curvature angle} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These figures are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This gives us a total of 9 image-derived parameters.

@@

\subsubsection{Assessing the Information Gain Using the PCA}
%%
To determine the information gain per image-derived variables, we calculated a PCA and compared the loadings per dimension.
%%
The first three principal components explain 75\% of variance in the image-derived variables.
%%
The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{curvature sagittal}, \emph{curvature angle} and \emph{curvature angle sagittal}.
%%
The second component, adding 16\% of variance, weigths  \emph{curvature coronal} and \emph{curvature angle coronal}.
%%
The third component explains 12\% of variance and weights \emph{torsion} and \emph{curvature transverse}.
%%
This supports our prior conclusion about the low information gain of the transverse planes.
%%
\emph{Torsion} also adds little variance to the information space.
%% ------------------------------------------
\subsubsection{Heterogenous Correlations}
%%
We then expanded our focus on correlations of image-derived parameters with all other non-image variables.
%%
We applied a heterogenous correlation technique to derive correlations between all variables is the data set.
%%
The method uses the following correlation metrics for the different type combination:
%%
\begin{itemize}
%%
\item \emph{pearson product-moment} for two continuous variables, 
\item \emph{polyserial} correlation for one continuous and one categorical variable, and
\item \emph{polychoric} correlations for two categorical variables.
\end{itemize}
\com{TODO: Abbildung einer Korrelationsmatrix? Waere sehr gross!}
%%
All correlation values are scaled between 0 - no correlation and 1 - identical.
%%
Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
%%
We display the resulting \emph{contingency matrix} using a heat map, mapping correlation values to color brightness with white for 0 and dark blue for 1 \cite{Klemm2014VIS}.
%%
We calculated the contingency matrix for all size groups and looked for correlations between image- and non-image derived variables.
%%
The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the parameter.
%%
Only weak correlations could be found for \emph{Mean Curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
%%
One surprising result was the small correlation with \emph{torsion}, which correlated with almost no variables (p values between 0 and 0.05) and \emph{parkinson} (0.24).
%%
This observation brought us to the conclusion to incorporate more sophisticated data mining techniques to assess the influence of the image-derived parameters.

@@

\paragraph{The Error Term. }
Normally the error rate for a classification is calculated with $error = \frac{n_{classifiedCorrectly}} {n}$ where $n$ is the number of subjects.
%%
The metric usually works well for variables with uniform distribution.
%%
It distorts the result for other distribution types.
%%
If for example a variable indicating a disease is negative for 90\% of the subject and the classifier simply assigns all subjects to \emph{not ill} the error metric would yield an error of 10\%, even though it is very bad.
%%
Our error term therefore incorporates the discriminative power of each manifestation and is denoted as follows:
%%
%Therefore the discriminative power of each dimension in the equation
%error = 1 - \frac{\[ \sum_{m=0}^M \] \frac{M_{classifiedCorrectly}}{M_all}}{number of dimensions}
\begin{equation}
error = 1 - \frac{\sum_{m=0}^M \frac{m_{classifiedCorrectly}}{m_{all}}}{number_{dimensions}}
\end{equation}
$M$ represents the set of manifestation of each variable.
%%
The error is scaled to denote perfect classification with 0 and 1 is equal to random selection.
% The error rate has to be judged depending on the number of classes of the target variable.
% %%
% An error rate of 50\% for a target variable with two manifestations makes the classifier equal to throwing a coin.
% %%
% The same error rate is acceptable a target variable with more than two manifestations.
% %%
% Picking a value for a variable of four manifestations with an error rate of 50\% is better than the random selection (error rate of 75\%).
% %%
% The following scales the error rate so that 0 denotes perfect classification and 1 is equal to random selection.
% %%
% \begin{equation}
% error_{normalized} = \frac{error}{100 - \frac{100}{class number}}
% \end{equation}
%%
It allows for comparability of error rates between variables with different manifestation count.

@@

%\com{Potentially skip these due to the confounding of body size}
\subsubsection{Age Groups}
%%
Gender could be described for each age group using \emph{mean curvature} and \emph{curvature angle}.
%%
The accuracy varies between $0,35$ ($43,8-54$ years of age, 3 decision rules) to $0,27$ ($21-43,8$ years of age, 6 decision rules).
%%
The starting age of smoking could be characterized well with an accuracy between $0,25$ to $0,3$ for all age groups except of $21,1-43,8$ years of age.
%%
The result however is probably over fitted to the data due to tree sizes between $16$ and $23$.
%%
\com{Feedback der Greifswalder notwendig, warum das so sein koennte.}
%%

Many variables, such as body size can be described with an error rate of $0,35$ to $0,45$ but only using large decision trees with over 20 rules.
%Body-size can be determined with an error rate of $0,36$ to $0,39$ for subject
% age groups, but also imposes over fitting to the data yielding a decision tree of over 30 rules.
%%
Notable is also the increasing accuracy with increasing decision tree size.
%%
For subjects between $43,8-54$ years of age, the classifier discriminated subjects with thyroid nodules with an error rate of $0,32$.

The dummy variable analysis shows many results using a decision tree with one rule based on \emph{mean curvature} with accuracy between $0,4$ and $0,5$.
%%
It shows that mean curvature can be used to predict dependencies to variables, such as \emph{high blood pressure}, \emph{hypothyroidism} or \emph{nutrition}.

\subsubsection{Size Groups}
%%
Many previously described results can be associated to be confounded with subject size.
%%
Differences in the gender analysis are mostly due to the average height difference between males and females.
%%
For example, large subjects are already characterized by their rather straight spine.
%
The question is, whether the inter-group spine-variability parameter is enough for predicting other parameter or not.
%%
Dividing subjects into height groups potentially highlights classifications not confounded by body-height.
%%
\paragraph{Large Decision Trees. }
Back-pain associated variables can be explained for various size-groups, but we could not extract universal rules.
%%
Radiating back pain could be described with error rate of $0,39$ using 23 rules for subjects between $171-177 cm$ body size using torsion and mean-curvature.
%%
For subjects sized $177-202 cm$ the accuracy drops to $0,47$ using 20  decision rules.
%%
There are several decision trees for laboratory values, such as alanine aminotransferase value in the blood can be described with a high precision of $0,4$ ($139-164 cm$) to $0,36$ ($164 - 171 cm$).
%%
Similar values can be observed for cholesterol or age.
%%
Due to the large decision trees, these results are not usable and impose overfitting to the data.
%%
\paragraph{Small Decision Trees. }
%%
The dummy variables show several variables described using only one decision rule with accuracy between $0,42$ to $0,47$.
%%
Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
%%
These variables include nutrition parameter, thyroid disorder and social problems induced by back pain.

@@

Another observation during the analysis are very large decision trees achieving low error scores.
%%
Their complexity impose overfitting to the data and they are not suitable for extracting universal rules.
%%
\subsection{Proposed Features}
\label{sec:ProposedFeatures}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.5\textwidth]{figures/placeholder}
  \caption{Proposed features. (a) dented vertebraes are a sign of heavy stress and bone absorption. (b) Spine canal thickness is associated with herniated disks. (c) the spine canal shape can be used to characterize scoliosis and lordosis as well as the \emph{Cobb-Angles} \cite{Cobb}. }
  \label{fig:proposedFeatures}
\end{figure}
%%
\noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain \com{Zitat Frau Hegenscheid fragen}.
%%
We have to capture earlier signs of pathological change in order to derive better predictive models.
%%

>>>>

\subsection{Image Data}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.5\textwidth]{figures/centerline}
  \caption{
	(a) left: Tetrahedron-based finite element model (FEM) of the lumbar spine, capturing the L1-L5 vertebrae as well as the lumbar spine canal.
	%%
	(a) right: FEM aligned to a MRI data set.
	%%
	The purple dashed line represents the centerline, describing the lumbar spine canal with 92 points.
	%%
	(b) Using the Frenet formulas, we extract the weighted sum of curvature and torsion for all 92 points (white dashes).
	%%
	The curvature angle ($\alpha$) is calculated using the top, bottom and middle point.
	%%
	We also extracted the variables for each projection axis which to assess their information gain.
	%(b) Using the Frenet formulas, we extract the curvature, torsion and the curvature angle from the lumbar spine canal center line.
	}
  \label{fig:centerline}
\end{figure}
\com{Image Data}
Four trained technicians acquired the medical image data for each subject in a standardized way using a 1.5~Tesla MRI scanner (Magnetom Avanto; Siemens Medical Solutions, Erlangen, Germany).
%%
The spine protocol consisted of a sagittal T1-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) and a sagittal T2-weighted turbo-spin-echo sequence ($1.1\times1.1\times4.0~mm$ voxels) \cite{Hegenscheid2013}.
%%

@@

We have to assess the model accuracy to extract key figures from it.
%%
The detection model depicts the vertebrae positions, but lacks detailed information about their volume.
%%
It captures reliable information about spine canal curvature.
%%
In a previous work \cite{Klemm2013VMV,Klemm2014VIS}, we extracted a centerline representation of the lumbar spine canal from the detection model (Fig.~\ref{fig:centerline}~a).
%%
Using the Frenet formulas \cite{Frenet}, we extracted the following metrics from the model (Fig.~\ref{fig:centerline}~b):
%%
%Thus, we extracted the following metrics from the model:
%%
%\paragraph{Curvature}
\begin{itemize}
	% \item The curvature of the centerline is calculated using the summed mean curvature between adjacent points:
% %%
% $\sum \frac{curvature_i}{curvature_all}$.
	\item \emph{Curvature} is calculated as weighted sum of curvature between all adjacent points describing the centerline: $\sum \frac{curvature_i}{curvature_{all}}$.
%%
	\item \emph{Torsion} (deviation of a curve from its current course) is calculated as weighted sum of torsion between all adjacent points describing the centerline: $\sum \frac{torsion_i}{torsion_{all}}$.
	\item \emph{Curvature angle $\alpha$} is defined by the middle point of the spine canal centerline as \emph{vertex} and the line between middle point and top/bottom point as \emph{sides}.
%%
\end{itemize}
%%
These figures are also extracted in the sagittal, coronal and transversal projection of the model.
%%
We assess the information gain of each dimension using these projections.
%%
This gives us a total of 9 image-derived parameters.

@@

\subsubsection{Assessing the Information Gain Using the PCA}
%%
To determine the information gain per image-derived variables, we calculated a PCA and compared the loadings per dimension.
%%
The first three principal components explain 75\% of variance in the image-derived variables.
%%
The first principal component explains 47\% of the variance and weights primarily \emph{mean curvature}, \emph{curvature sagittal}, \emph{curvature angle} and \emph{curvature angle sagittal}.
%%
The second component, adding 16\% of variance, weigths  \emph{curvature coronal} and \emph{curvature angle coronal}.
%%
The third component explains 12\% of variance and weights \emph{torsion} and \emph{curvature transverse}.
%%
This supports our prior conclusion about the low information gain of the transverse planes.
%%
\emph{Torsion} also adds little variance to the information space.
%% ------------------------------------------
\subsubsection{Heterogenous Correlations}
%%
We then expanded our focus on correlations of image-derived parameters with all other non-image variables.
%%
We applied a heterogenous correlation technique to derive correlations between all variables is the data set.
%%
The method uses the following correlation metrics for the different type combination:
%%
\begin{itemize}
%%
\item \emph{Pearson product-moment} for two continuous variables, 
\item \emph{Polyserial} correlation for one continuous and one categorical variable, and
\item \emph{Polychoric} correlations for two categorical variables.
\end{itemize}
\com{TODO: Abbildung einer Korrelationsmatrix? Waere sehr gross!}
%%
All correlation values are scaled between 0 - no correlation and 1 - identical.
%%
Some variables are too sparse for calculating correlations, for example \emph{treatment of diabetes}, or \emph{medication against high blood pressure}.
%%
We display the resulting \emph{contingency matrix} using a heat map, mapping correlation values to color brightness with white for 0 and dark blue for 1 \cite{Klemm2014VIS}.
%%
We calculated the contingency matrix for all size groups and looked for correlations between image- and non-image derived variables.
%%
The resulting contingency matrices are available as supplementary material and show no strong correlation with any of the parameter.
%%
Only weak correlations could be found for \emph{Mean Curvature} with \emph{gender} (0.42), \emph{body size} (0.39) and \emph{number of born children} (0.29).
%%
One surprising result was the small correlation with \emph{torsion}, which correlated with almost no variables (p values between 0 and 0.05) and \emph{parkinson} (0.24).
%%
This observation brought us to the conclusion to incorporate more sophisticated data mining techniques to assess the influence of the image-derived parameters.

@@

\paragraph{The Error Term. }
Normally the error rate for a classification is calculated with $error = \frac{classifiedCorrectly_{n}} {n}$ where $n$ is the number of subjects.
%%
The metric usually works well for variables with uniform distribution.
%%
It distorts the result for other distribution types.
%%
If for example a variable indicating a disease is negative for 90\% of the subject and the classifier simply assigns all subjects to \emph{not ill} the error metric would yield an error of 10\%, even though it is very bad.
%%
Our error term therefore incorporates the discriminative power of each manifestation and is denoted as follows:
%%
%Therefore the discriminative power of each dimension in the equation
%error = 1 - \frac{\[ \sum_{m=0}^M \] \frac{M_{classifiedCorrectly}}{M_all}}{number of dimensions}
\begin{equation}
error = 1 - \frac{\sum_{m=0}^M \frac{classifiedCorrectly_{m}}{all_{m}}}{number_{dimensions}}
\end{equation}
$M$ represents the set of manifestation of each variable.
%%
The error is scaled to denote perfect classification with 0 and 1 is equal to random selection.
% The error rate has to be judged depending on the number of classes of the target variable.
% %%
% An error rate of 50\% for a target variable with two manifestations makes the classifier equal to throwing a coin.
% %%
% The same error rate is acceptable a target variable with more than two manifestations.
% %%
% Picking a value for a variable of four manifestations with an error rate of 50\% is better than the random selection (error rate of 75\%).
% %%
% The following scales the error rate so that 0 denotes perfect classification and 1 is equal to random selection.
% %%
% \begin{equation}
% error_{normalized} = \frac{error}{100 - \frac{100}{class number}}
% \end{equation}
%%
It allows for comparability of error rates between variables with different manifestation count.

@@

%\com{Potentially skip these due to the confounding of body size}
\subsubsection{Age Groups}
%%
Gender could be described for each age group using \emph{mean curvature} and \emph{curvature angle}.
%%
The accuracy varies between $0.35$ ($43.8-54$ years of age, 3 decision rules) to $0.27$ ($21-43.8$ years of age, 6 decision rules).
%%
The starting age of smoking could be characterized well with an accuracy between $0.25$ to $0.3$ for all age groups except of $21.1-43.8$ years of age.
%%
The result however is probably over fitted to the data due to tree sizes between $16$ and $23$.
%%
\com{Feedback der Greifswalder notwendig, warum das so sein koennte.}
%%

Many variables, such as body size can be described with an error rate of $0.35$ to $0.45$ but only using large decision trees with over 20 rules.
%Body-size can be determined with an error rate of $0,36$ to $0,39$ for subject
% age groups, but also imposes over fitting to the data yielding a decision tree of over 30 rules.
%%
Notable is also the increasing accuracy with increasing decision tree size.
%%
For subjects between $43.8-54$ years of age, the classifier discriminated subjects with thyroid nodules with an error rate of $0.32$.

The dummy variable analysis shows many results using a decision tree with one rule based on \emph{mean curvature} with accuracy between $0.4$ and $0.5$.
%%
It shows that mean curvature can be used to predict dependencies to variables, such as \emph{high blood pressure}, \emph{hypothyroidism} or \emph{nutrition}.

\subsubsection{Size Groups}
%%
Many previously described results can be associated to be confounded with subject size.
%%
Differences in the gender analysis are mostly due to the average height difference between males and females.
%%
For example, large subjects are already characterized by their rather straight spine.
%
The question is, whether the inter-group spine-variability parameter is enough for predicting other parameter or not.
%%
Dividing subjects into height groups potentially highlights classifications not confounded by body-height.
%%
\paragraph{Large Decision Trees. }
Back-pain associated variables can be explained for various size-groups, but we could not extract universal rules.
%%
Radiating back pain could be described with error rate of $0.39$ using 23 rules for subjects between $171-177~cm$ body size using torsion and mean-curvature.
%%
For subjects sized $177-202~cm$ the accuracy drops to $0.47$ using 20  decision rules.
%%
There are several decision trees for laboratory values, such as alanine aminotransferase value in the blood can be described with a high precision of $0.4$ ($139-164~cm$) to $0.36$ ($164 - 171~cm$).
%%
Similar values can be observed for cholesterol or age.
%%
Due to the large decision trees, these results are not usable and impose overfitting to the data.
%%
\paragraph{Small Decision Trees. }
%%
The dummy variables show several variables described using only one decision rule with accuracy between $0,42$ to $0,47$.
%%
Most of these variables have a dominant manifestation and the classifier shows a low detection precision for the second manifestation.
%%
These variables include nutrition parameter, thyroid disorder and social problems induced by back pain.

@@

Another observation during the analysis are very large decision trees achieving low error scores.
%%
Their complexity impose overfitting to the data and they are not suitable for extracting universal rules.
%%
\subsection{Proposed Features}
\label{sec:ProposedFeatures}
\begin{figure}[!h]
  %\vspace{-0.2cm}
  \centering
  \includegraphics[width=0.5\textwidth]{figures/placeholder}
  \caption{
	Proposed features.
	%%
	(a) dented vertebraes are a sign of heavy stress and bone absorption.
	%%
	(b) Spine canal thickness is associated with herniated disks.
	%%
	(c) the spine canal shape can be used to characterize scoliosis and lordosis as well as the \emph{Cobb-Angles} \cite{Cobb}.
	%%
	}
  \label{fig:proposedFeatures}
\end{figure}
%%
\noindent Pathological deformation of the lumbar spine is usually the last stage and is associated with very strong lower back pain \com{Zitat Frau Hegenscheid fragen}.
%%
We have to capture earlier signs of pathological change in order to derive better predictive models.
%%
